Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-11-11 AnalyticStructure_main.tex
Further modifications. Almoste ready to insert larger molecules in the manuscript

% Useful packages
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}
\usepackage[markup=underlined]{changes}
\definechangesauthor{dal}
\definechangesauthor{gen}

@@

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1]{\op{H_1}\argument{#1}}
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}
\newcommand{\fscd}[1]{\ket{f_p^\Phi(\omega)^{#1}}}
\newcommand{\GH}{\op G_{\op T}}

@@

\section{Introduction}

% Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the
% computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
% For example, it may be in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
% The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
% other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
% Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
% that a given computational model provides for a specific set of quantities, regardless of their potential agreement with
% experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
% of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
% employing the same model can be compared.
% 
% A recent, remarkable example of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
% where a grassroot community of code-owners struggled to extract well-defined computational results in the context of
% Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as lattice constants and bulk modulus of
% elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties
% on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
% Any other code able to extract the same quantities with the same approximations may be compared with the
% results of the community in order to assess its computational reliability.
% 
% For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
% reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
% model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
% only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
% For communities like DFT, such a ``calibration'' of computer codes among each other is a discipline that is still in its infancy;
% the DeltaCodes initiative and the foundations of DFT are separated in time by more than 50 years, and the computational
% model employed in the comparison (the Perdew-Burke-Erzerhof functional~\cite{PBE}) has been introduced more than 20 years ago.

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
For communities like DFT, the need of such a ``calibration'' of computer codes has led to the DeltaCodes project~\cite{deltaTest2016}. 
This project represents a remarkable example of precision-driven initiative where a grassroot community of code-owners struggled to extract 
well-defined computational results in the context of Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities 
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
In order to do this, we perform analytic considerations on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others that would \emph{explicitly} depend of the employed computational treatment.

@@

\replaced[id=dal]{Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities. In particular we revisit the impact 
that the Isolated BC have on the solutions of the time-independent Schr\"odinger Equation.}{Let us start our discussion by revisiting the 
impact that the Isolated BC have on the solutions of the time-independent Schr\"odinger Equation. We here present well-known concepts of 
quantum mechanics, under a perspective that will turn out to be very useful in the forthcoming considerations about linear-response quantities.}

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states}
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
\added[id=dal]{Fluctuation states are complex quantities and consequently $\dm'_\Phi(\omega)$ is not hermitian; nonetheless the $\omega$ dependence of \eqref{rhoPrimeFluctuationStateDef1} ensures that the hermitian nature of this operator is recovered in the time domain, as expected.}

An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains,
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \GH(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op \Phi(\omega) +\op V'[\dm'_\Phi](\omega) )\ket{\psi_p}
$ as a source term.

\replaced[id=dal]{Equation \eqref{rhoPrimeFluctuationStateDef2} is the convenient starting point to assess the localization properties of the FS as it is easy to see that, for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.}{It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level
that governs the localization properties of the FS.}
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is \added[id=dal]{spatially} localized, \textit{i.e.} its evaluation can be restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

\added[id=dal]{Added a subsection for static perturbations}

\subsection{Localized behavior for static perturbations}

The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strenghts $F_j$ small enough to
preseve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

\added[id=dal]{Modified the name of the subsection}

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that \added[id=dal]{(equation for the response density modified)}
% \be\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = 2 \mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
% \ee
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O} = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
Let us suppose that also the observable of interest $\op O$ exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.

We observe that there exists two
distinct regimes, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator.
This is naturally achieved in the above mentioned Krylov space treatments, which
explicitly deal with the expression of the scalar products. However
it is important that the underlying basis set is able to express delocalised states.

@@

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors.
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues.
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behave as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$
converges to its reference value.

The second sector contains a continuum spectrum of delocalized excitations. The associated states $\ket{\phi_p^a}$ behave as the unbound virtual orbitals analyzed in \cite{boffi2016}
and are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and
the associated spectrum exhibits an explicit dependence from the size of the domain.
Excitations belonging to this sector thus lose their intrinsic physical meaning and have to be understood simply as
a representation of the density of states (DOS) of the Liouvillian superoperator.

@@

\subsection{Physical relevance of the continuum sector}

We have presented formal arguments, confirmed by numerical calculations, that show that
the excitations belonging to the discrete sector of the spectrum have energies
that can be \emph{quantitatively} compared between codes, thanks to their
localized character.
Such objects are LR quantities, associated to the molecule, whose numerical convergence
can be found by employing techniques which can be borrowed from the ones adopted for GS calculations. Being expressed by localized excited states,
computational setups with sufficient level of completeness are able to
express with high \emph{precision} such section of the excitation spectrum.

>>>>

% Useful packages
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{tikz}
\usepackage{xifthen}
%\pgfplotsset{compat=1.15}
% \usepackage[markup=underline]{changes}
\usepackage[final]{changes}
\definechangesauthor{dal}
\definechangesauthor{gen}

@@

% special operators
\newcommand{\dmnot}{\op{\rho}_0}
\newcommand{\dm}{\op{\rho}}
\newcommand{\hnot}{\op{H}_0}
\newcommand{\hone}[1]{\op{H}_1\argument{#1}}
\newcommand{\transition}[1]{\op T_{#1}}
\newcommand{\excite}[2]{\op e_{{#1}{#2}}}
\newcommand{\decay}[2]{\op d_{{#1}{#2}}}
\newcommand{\Liouv}{\sop{\mathcal L}}
\newcommand{\Liouvnot}{\sop{\mathcal L_0}}
\newcommand{\coupl}{\sop{\mathcal K}}
\newcommand{\honetmp}[1]{\op{H_1}\argument{#1}}
\newcommand{\identity}{\op{\mathbb I}}
\newcommand{\rmat}[1]{\optr R}
\newcommand{\fscd}[1]{\mathrm{Im}\ket{f_p^{\mathbf r_j}(\omega)^{#1}}}
\newcommand{\GH}{\op G_{\op T}}

@@

\section{Introduction}
\deleted[id=gen]{%
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the
computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
For example, it may be in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.
%
A recent, remarkable example of precision-driven initiative is provided by the DeltaCodes project~\cite{deltaTest2016},
where a grassroot community of code-owners struggled to extract well-defined computational results in the context of
Density Functional Theory (DFT) calculations. Ground-State quantum mechanical quantities as lattice constants and bulk modulus of
elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties
on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.
%
For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
For communities like DFT, such a ``calibration'' of computer codes among each other is a discipline that is still in its infancy;
the DeltaCodes initiative and the foundations of DFT are separated in time by more than 50 years, and the computational
model employed in the comparison (the Perdew-Burke-Erzerhof functional~\cite{PBE}) has been introduced more than 20 years ago.}

For a theoretical approach that is based on numerical calculations, for which no analytic reference solution exists,
reducing the computational uncertainty is the only possible way to shed light on the predictivity power of the
model. In other terms, the ``accuracy'' of a result with respect to experimental data may be reliably quantified
only when the computational uncertainty is guaranteed to be significantly lower than the found discrepancy.
For communities employing Density Functional Theory (DFT), the need of such a ``calibration'' of computer codes has led to the DeltaCodes project~\cite{deltaTest2016}.
This project represents a remarkable example of precision-driven initiative where a grassroot community of code-owners struggled to extract
well-defined computational results in the context of DFT calculations. Ground-State quantum mechanical quantities
as lattice constants and bulk modulus of elementary crystals extracted from different codes were compared with each other, with the objective of reducing the uncertainties on such data, and in some sense, to define a \emph{calibration} procedure for a Solid-State Physics DFT code.
Any other code able to extract the same quantities with the same approximations may be compared with the
results of the community in order to assess its computational reliability.

@@

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems,
in particular referring to LR-TDDFT calculation. We will investigate the impact that
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
\deleted[id=gen]{In order to do this}We perform analytic considerations on the equations beyond LR treatment and discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects. Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - from others that would \emph{explicitly} depend of the employed computational treatment.

@@

\replaced[id=dal]{Let us start our discussion by reviewing some well-known concepts of quantum mechanic under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities. In particular we revisit the impact
that the Isolated BC have on the solutions of the Schr\"odinger Equation.}{Let us start our discussion by revisiting the
impact that the Isolated BC have on the solutions of the time-independent Schr\"odinger Equation. We here present well-known concepts of
quantum mechanics, under a perspective that will turn out to be very useful in the forthcoming considerations about linear-response quantities.}

@@

We can therefore introduce an explicit representation of the
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} (FS)
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$.
\added[id=gen]{
However, non-hermiticity of $\dm'_\Phi(\omega)$ implies that the
fluctuation states are complex quantities. Causality implies that the real and imaginary part of the states are related to each othe by the Kramers-Kronig relations}
\be\lb{KKrel}
\mathrm{Re}\left( \ket{f_p^\Phi(\omega)} \right) =
\mathcal P \int \dd \omega'
\frac{\mathrm{Im}\left(\ket{f_p^\Phi(\omega')}\right)}{\omega -\omega'}.
\ee
\added[id=gen]{We will come to these equations in the forthcoming sections.}
%}

An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states  are
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains,
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be recast in a scattering problem:
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \GH(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op \Phi(\omega) +\op V'[\dm'_\Phi](\omega) )\ket{\psi_p}
$ as a source term.

\replaced[id=dal]{Equation \eqref{rhoPrimeFluctuationStateDef2} is the convenient starting point to assess the localization properties of the FS as it is easy to see that, for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level that governs the asymptotic behavior of the associated Helmholtz kernel.}{It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level
that governs the localization properties of the FS.}
As a ulterior \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is \added[id=dal]{spatially} localized, \textit{i.e.} its evaluation can be restricted on a bounded real-space domain.
Even when such locality argument holds true, for $\omega > |\eps_p|$,
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range,
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.
\added[id=gen]{These arguments demonstate that for frequencies above the IP threshold energy, the
response density operator cannot be expressed in terms of localized states. Locality of
FS (and therefore of $\dm'_\Phi(\omega)$) can be only recovered for frequencies below such threshold.}

\deleted[id=dal]{Added a subsection for static perturbations}

\subsection{Localized behavior for static perturbations}
The locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain.
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem.
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques,
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
\begin{figure}[t]
\includegraphics[scale=0.68]{Fig1_CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line)
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest
box.}
\end{figure}

@@

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} =
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
where $\mathbf d^\text{(GS)}=\braket{\mathbf r}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule
perturbed by a static electric field.
Care has been taken in keeping the field strenghts $F_j$ small enough to
preseve the validity of the linear response regime.
The local character of the fluctuation states at $\omega=0$ is
probed by verifying the convergence of the statical polarizability versus
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the
one of the static dipole of the molecule computed at zero external field.
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

\deleted[id=dal]{Modified the name of the subsection}

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$,
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response
density in terms of FS implies that \deleted[id=dal]{(equation for the response density modified)}
% \be\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}_\Phi(\omega) = 2 \mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
% \ee
\begin{multline}\lb{LinearResponseFunctDef2}
\braket{\delta\op O} = \sum_p \left( \bra{f_p^\Phi(-\omega)}\op O\ket{\psi_p}+ \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \\
=2 \sum_p\bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}  \;.
\end{multline}
Let us suppose that also the observable of interest $\op O$ \replaced[id=gen]{may be restricted to localized domains in the same way as the potentials $ \op V$ and $\op \Phi$.}{exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.}
Once again, it would be enough to consider a local operator for this condition to be met.

Two
distinct regimes can be identified, on the basis of value of $\omega$.
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2}
is expressed in terms of genuinely localized
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.
 \added[id=gen]{In this regime, a computational treatment based on localized basis set is susceptible to provide a precise answer, assuming a reasonable level of completeness.}

On the contrary, the computation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation
states start behaving as unbound wave functions.
As by hypothesis $\ket{\op O \psi_p}$ is bound-state like, \emph{only} the scalar products of Eq.~\eqref{LinearResponseFunctDef2} can be evaluated in a localized domain.
In this case a computational description of fluctuation states is only meaningful in ``weak'' sense,
where the bound state character of $\{\bra{\psi_p}\op O\}$ behaves as a regulator.
This is naturally achieved in the above mentioned Krylov space treatments, which
explicitly deal with the expression of the scalar products.
\added[id=gen]{However, in this regime, imposing \emph{by design} a localized
behaviour of the FS might reveal very dangerous in view of convergence of the results:
the computational basis set have to be able to express in the support domain of $\{\bra{\psi_p}\op O\}$ a asymptotically delocalized FS. The degrees of completeness
of the computational basis should therefore be increased by imposing oscillatory degrees of freedom needed to represent the FS at the border of the simulation domain: it is important that the underlying basis set is able to express delocalised states.}

@@

\vspace{1cm}
\added[id=dal]{A proposal to reorganize the material of the following two subsection B e C in a different way}

\section{Physical relevance of the excitation spectrum. A comparison among localized and continuum sectors}
\deleted[id=dal]{
Comments on the features and the physical relevance of the two sector of the excitation spectrum}

%\subsection{Localized sector}

%\begin{itemize}
% \deleted[id=dal]{\item It is composed by a genuine set of discrete elements.}
% \deleted[id=dal]{\item The values of the excitation energies do not depend on the computational setup (observable features) and so can be compared among different approaches.}
% \deleted[id=dal]{\item The knowledge of this sector of the excitation spectrum allows us to express only the imaginary part of (proper equation that expresses the response functional written using Sokowsky-Plemely
% on the susceptibility functional).}
% \deleted[id=dal]{\item A computational setup in which delocalized degrees of freedom are excluded by design can in principle describe to this sector in a exact way.}
%\end{itemize}

\subsection{Continuum sector}

\begin{itemize}
 \deleted[id=dal]{\item Not true poles of the susceptibility (but only an effective tool to compute the spectral function).}
 \deleted[id=dal]{\item Deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and the associated spectrum exhibits an explicit dependence from the size of the domain.}
 \deleted[id=dal]{\item This sector of the excitation spectrum is needed to compute imaginary parts above IP and real part for \emph{all} the values of $\omega$. Add examples on the statical polarizability. Can we comment}
 \item Make some statements on the features that a computational setup must have to guarantees that this sector provides converged results. Difference between systematic and localized basis setup. Can we comment about the needs to add ghost atoms for the evaluation of statical polarizability with gaussian basis?
\end{itemize}

\vspace{1cm}

@@

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors.
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues.
Borrowing the terminology introduced in section \ref{SEopenSystem} to the present case, we can affirm that such localized excitations behave as
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$
converges to its reference value.
\added[id=gen]{A computational setup in which delocalized degrees of freedom are excluded by design can, in principle, describe excitations of this sector in a exact way.}

The second sector contains a continuum spectrum of delocalized excitations. The associated states $\ket{\phi_p^a}$ behave as the unbound virtual orbitals analyzed in \cite{boffi2016}
and are deeply affected by the computational setup used to represent them. In particular, the presence of a finite computational domain implies a fictitious discretization of the excitation levels and
the associated spectrum exhibits an explicit dependence from the size of the domain.
Excitations belonging to this sector thus lose their intrinsic physical meaning and have to be understood simply as objects providing an effective representation of the density of states (DOS) of the Liouvillian superoperator.

@@

\subsection{Relative importance of discrete and continuum excitations}
\deleted[id=gen]{
We have presented formal arguments, confirmed by numerical calculations, that show that
the excitations belonging to the discrete sector of the spectrum have energies
that can be \emph{quantitatively} compared between codes, thanks to their
localized character.
Such objects are LR quantities, associated to the molecule, whose numerical convergence
can be found by employing techniques which can be borrowed from the ones adopted for GS calculations. Being expressed by localized excited states,
computational setups with sufficient level of completeness are able to
express with high \emph{precision} such section of the excitation spectrum.}
